A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-22 graph_syntax.tex
stuff

\section{Graph Representation of MDAO Constructs}
\label{s:graph representation}
In this section we present our graph syntax applied to the description of MDAO 
Problems. We make use of the Sellar problem, described in 
Eqn. \ref{eqn:simple_fpf} as an example to illustrate the implementation of the syntax. 
The graph representation of the Sellar problem, using the new stynax, is 
presented in Fig. \ref{f:sellar_graph_full}. In this section sub-graphs or 
slight modifications of the full graph will be used to demonstrate the 
important MDAO constructs described by this graph syntax.

@@

  Note that the variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $D_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

@@

  A local analysis block is defined as a block with incomming edges from 
  exactly one other analysis block and outgoing edges from exactly one other analysis 
  block.

  A global analysis block is defined as

  Fig. 
  \ref{f:sellar_graph_full} has expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $D_1$ and 
  $D_2$, so it would considered global.

  Simply stating that any node or block with incomming or outgoing edges
  from more than one other node or block is global works glosses over a more 
  subtle aspect of some MDAO problems. The locality of any node in a graph is 
  a relative property. For instance, a single node might have outgoing edges 
  to two separate analysis blocks, $A$ and $B$, but none to a third, $C$. Then 
  you could say that this node was global with respect to $A$ and $B$ separately, but local 
  to the group $A,B$. This situation produces a natural 
  hierarchy in a graph, where you could form increasingly smaller groups as you 
  segment problem into finer localities.

  When solving a problem using a heirarchical decomposition approach, 
  there are various techniques for partitioning that are effective in different 
  scenarios \cite{krishnamachari1997optimal,michelena1997hypergraph,sobieszczanski1997,Perez2004,allison2009optimal}. 
  Tosserams et al proposed a language based syntax for describing the partitioning of problems, named $\Psi$
  \cite{tosserams2010specification}. $\Psi$ provides an opposing perspective to 
  this work, where they compose hierarchies from the bottom up, into larger and 
  larger groups. However, $\Psi$ provides a compiler that can produce a standard 
  problem representation of the assembled problem, and a number of example converters
  from that standard representation to application specific formats.

  [[Could probably build this and include it in appendix]]

@@

  Coupling exists in a design problem when a set of two or more analyses each depend on the
  others outputs. In the Sellar problem from Eqn \ref{eqn:simple_fpf} the two 
  coupling constraints, $H_1$ and $H_2$ provide a reciprical dependence between 
  $D_1$ and $D_2$. In a graph, these constraints show up as connection edges between 
  the outputs of each analysis block. The relevant connection edges are highlighted in 
  Fig. \ref{f:coupling}, where they form a cycle. Cycles of connection edges are the characteristic 
  structure that represents coupling.

@@

  It is important to note that the coupling cycles do not contain 
  any driver nodes. Solver, optimizer, or other iterative process is 
  invovled in the coupling definition--though one will be necessary for a valid 
  PSG. Additionally, a coupling cycle has no inherant start or end. It would be
  valid to pick any node in the cycle as a starting point, and proceed around the
  loop untill you get back to your starting point. For the Sellar Problem, picking 
  $D_1$ as the starting point would yeild a problem as given in 
  Eqn.~\ref{eqn:simple}, while picking $D_2$ would yield the problem as given in 
  Eqn.~\ref{eqn:simple2}.

  Larger problems can contain more complex cycles in their FPG, indicating more 
  complex coupling between analyses. A cycle can contain more than 
  just two analysis blocks. Multiple independent cycles can exist, indicating 
  independent coupling relationships. Cycles can also overlap, where the same analysis 
  blocks are involved in multiple different coupling cycles. All of these situations
  arise naturally as the size of problems grows, and managing this coupling may
  become difficult. In the present work, Sec.~\ref{s:example problem}.\ref{ss:obtaining an FPG} 
  will demonstrate how building an FPG from an MCG provides an opportunity to 
  identify and potentailly reduce the number of cycles in a graph.

  If large amounts of coupling are present it can help convergence 
  rates to look for an effective ordering for the execution of analyses.
  Rogers' DEMAID tool uses a genetic algorithm to find an ordering that minimizes 
  the overall coupling of the system by separating independent cycles in the 
  graph \cite{rogers1996,rogers1996demaid}. Rodgers work focused on the matrix 
  form of the DSM for ordering optimization. Lu and Martins more recent work 
  worked with a weighted form of the DSM and used an iterative clustering approach 
  to perform a similar task to DEMAID \cite{Lu2012}.

  However, when using a graph representation, the ordering problem can be greatly 
  simplified. For any given coupling cycle, a Gauss-Seidel iteration pattern 
  is automatically given simply folowing the cycle around the graph. If only 
  once cycle is present, then the task reduces down to selecting a starting point
  in the cycle. When sub-cycles are present each sub-cycle ordering is also 
  given automatically by following the graph, but now a there is freedom at which 
  level to manage the iteration over the cycles. Ordering of many cycles and 
  sub-cycles is still a challenge, but the total search space is reduced by 
  always following the ordering from the graph from a selected starting node.

\subsection{Multi--fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi--fidelity problem can be characterized by two or more different analyses 
  each calculating the same data. Mapping this type of multi--fidelity situation 
  onto a graph will yield two analysis blocks, each with their own output 
  variable nodes, having outgoing edges to a single variable in a third analysis 
  block that needs the data as input. Figure \ref{f:collision_example} shows 
  a modified version of the Sellar problem with a new analysis, $D_0$, representing a 
  low fidelity version of $D_1$.

  \begin{figure}
    \begin{center}
      %\includegraphics[height=.25\textheight]{}
    [INSERT GRAPH HERE]
    \caption{Graph of a notional multi-fidelity problem \label{f:collision_example}}
  \end{center}
  \end{figure}

  When an input variable node has more than one incoming edge, designers must make
  some kind of decision about how to manage the flow of information in their problem. 
  In Eqn.~\ref{e:collision} we defined a collision as the situation where
  there are more incoming connection edges than the upper indegree limit. So desginers can
  indicate how to handle conflicting edges in a problem graph by setting the upper indegree limit
  for a given node. In more concrete terms, setting this limit equates to defining 
  the part of the graph in question as single or multi--fidelity.

  When $\txt{deg}_u^-(y)=1$, then any conflicting edges in the graph will cause a collision
  and a choice between the colliding analysis blocks must be made. This results in a 
  single fidelity design problem. On the other hand if $\txt{deg}_u^-(y)=2$, then two 
  different analysis blocks can co-exist with a given graph without causing a 
  collision. The design problem is then a multi-fidelity problem.

  Multi-fidelity problems are always charaterized by graphs with variable 
  nodes that have an upper-degree-limit greater than one. These problems require 
  special techniques for resolving the confliciting edges by introducing some mechanism
  to manage when each of the different fidelity analyses are 
  run\cite{march2012provably,alexandrov2001approximation,Huang_Allen_Notz_Miller_2006}.

>>>>

\section{Graph Representation of MDAO Constructs}
\label{s:graph representation}
In this section we present our graph syntax applied to the description of MDAO 
Problems. We make use of the Sellar problem, described in 
Eq. \ref{eqn:simple_fpf} as an example to illustrate the implementation of the syntax. 
The graph representation of the Sellar problem, using the new stynax, is 
presented in Fig. \ref{f:sellar_graph_full}. In this section sub-graphs or 
slight modifications of the full graph will be used to demonstrate the 
important MDAO constructs described by this graph syntax.

@@

  The variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $D_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

@@

  A local analysis block is 
  defined as a block with incomming edges from exactly one other analysis 
  block and outgoing edges from exactly one other analysis block. A global 
  analysis block is defined as a block with incomming edges or outgoing edges 
  to more than one analysis block.

  [[MATH HERE]]

  Looking at the Sellar problem in 
  Fig. \ref{f:sellar_graph_full} we see the expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $D_1$ and 
  $D_2$, so it would considered global.

  Catagorizing any node or analysis block block with incomming or outgoing edges
  from more than one source as global neglects a more 
  subtle aspect of some MDAO problems. Namley the locality of any node in a graph is 
  a relative property. For instance, a single node might have outgoing edges 
  to two separate analysis blocks, $A$ and $B$, but none to a third, $C$. This 
  node was global with respect to $A$ and $B$ separately, but local 
  to the group $A,B$. This situation produces a natural hierarchy in a graph, 
  from which one could form ever smaller groups as the problem is segmented into 
  finer localities.

  When solving a problem using a hierarchical decomposition approach, 
  there are various techniques for partitioning a problem into hierarchical 
  localities that are effective in different scenarios \cite{krishnamachari1997optimal,michelena1997hypergraph,sobieszczanski1997,Perez2004,allison2009optimal}. 
  Tosserams \textit{et al.} proposed a language based-syntax for describing the 
  partitioning of problems, named $\Psi$\cite{tosserams2010specification}. 
  $\Psi$ provides an opposing perspective to this work, where they compose 
  hierarchies from the bottom up, into larger and larger groups. However, 
  $\Psi$ provides a compiler that can produce a standard problem representation 
  of the assembled problem and a number of example converters from that 
  standard representation to application specific formats.

@@

  Coupling exists in a design problem when a set of two or more analyses each depend on the
  other's outputs. In the Sellar problem from Eqn \ref{eqn:simple_fpf} the two 
  coupling constraints, $H_1$ and $H_2$, provide a reciprical dependence between 
  $D_1$ and $D_2$. In a graph, these constraints appear as connection edges between 
  the outputs of each analysis block. The relevant connection edges are highlighted in 
  Fig. \ref{f:coupling}, where they form a closed path between two analysis blocks,
  called a cycle. Cycles of connection edges are the characteristic structure that 
  represents coupling.

@@

  Coupling cycles do not contain driver nodes in the MCG of FPG representation. 
  Solvers, optimizers, and other iterative processes are not invovled in the coupling 
  definition--though one will be necessary to build a valid PSG from a given FPG. 
  Additionally, a coupling cycle has no inherent start or end. It would be acceptable to select
  node in the cycle as a starting point and proceed around the
  loop until you get back to the starting point. For the Sellar Problem, selecting 
  $D_1$ as the starting point would yield a problem as given in 
  Eq.~\ref{eqn:simple}, whereas selecting $D_2$ would yield the problem as given in 
  Eq.~\ref{eqn:simple2}.

  Larger problems can contain more complex cycles in their FPG, indicating more 
  complex coupling between analyses. For example A cycle can involve more than 
  just two analysis blocks. Multiple independent cycles could also exist, indicating 
  independent coupling relationships. Cycles can also overlap, meaning that the same analysis 
  blocks are involved in multiple different coupling cycles. All of these situations
  arise naturally as the size of problems grows, and managing this coupling may
  become difficult. In the present work, Sec.~\ref{s:example problem}.\ref{ss:obtaining an FPG} 
  demonstrates how building an FPG from an MCG provides an opportunity to 
  identify and potentailly reduce the number of cycles in a graph.

  If many couplings are present, convergence rates can be improved by 
  searching for an effective ordering for the execution of analyses.
  Rogers' DEMAID tool uses a genetic algorithm to find an ordering that minimizes 
  the overall coupling of the system by separating independent cycles in the 
  graph \cite{rogers1996,rogers1996demaid}. Rogers work focused on the matrix 
  form of the DSM for ordering optimization. Lu and Martins more recently leveraged 
  a weighted form of the DSM and used an iterative clustering approach to perform a 
  similar task to DEMAID \cite{Lu2012}.

\subsection{Multi--fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi--fidelity problem can be characterized by two or more different analyses 
  each calculating the same data. Representing this type of multi--fidelity 
  construct in a graph will yield two analysis blocks, each with its own output 
  variable nodes, having outgoing edges that connect to a single variable in a 
  third analysis block that needs the data as input. So the set of all multi-fidelity
  variables, $V_{mf}$, is given as:

  \begin{equation}
    V_{mf} = \{ V \st \txt{deg}^-(v)>1 \} \cup \{ V \st \txt{deg}^-_u(v)>=\txt{deg}^-(v) \}
  \end{equation}

  In a multi-fidelity problem, a given variable node may have multiple incomming 
  edges without causing a collision as defined by by Eq.~\ref{e:collision}. 
  Fig.~\ref{f:collision_example} shows a modified version of the Sellar problem 
  with a new analysis, $D_0$, representing a low fidelity version of $D_1$. 
  In this modified graph, the annotations indicate the upper indegree limit. For 
  $D_2$, $\txt{deg}^-_u(D_2.y_1) = 2$ which avoids a collsion. A  
  $\txt{deg}^-_u(v) > 1$ for any variable in a graph represents a decision to 
  allow multiple fidelities to interact at that part of the graph in order to alieviate
  a conflict. In section \ref{ss:obtaining FPG} we present an algorithm for finding 
  conflicts within a given MCG so that a designer can make the necessary decisisons about each 
  one in turn.

  \begin{figure}
    \begin{center}
      %\includegraphics[height=.25\textheight]{}
    [INSERT GRAPH HERE]
    \caption{Graph of the modified Sellar, multi--fidelity problem with $\txt{deg}^-_u(v)$ annotations on variable nodes\label{f:collision_example}}
  \end{center}
  \end{figure}

  Multi--fidelity problems are always charaterized by graphs with variable 
  nodes that have an $\txt{deg}^-_u(v) > 1$. These problems require 
  special techniques for resolving the confliciting edges by introducing some mechanism
  to manage when each of the different fidelity analyses are 
  run\cite{march2012provably,alexandrov2001approximation,Huang_Allen_Notz_Miller_2006}.
  The specifics of this mechanism are not given in any form withing an MCG or 
  FPG. Instead the multi-fidelity mechanism specifics would be represented as a 
  driver in a PSG derived from any given multi-fidelity FPG.
